---
title: "Matrix Algebra Review"
author: "Eric Zivot"
date: "Wednesday, June 30, 2016"
output: slidy_presentation
---

```{r echo=FALSE}
options(digits=3)
```

## Create Matrix Object

You create matrices using the <code>matrix()</code> constructor function:

```{r}
args(matrix)
# fill matrix by column (default)
matA = matrix(data=c(1,2,3,4,5,6), nrow=2, ncol=3)
matA
class(matA)
# fill matrix by row
matA = matrix(data=c(1,2,3,4,5,6), nrow=2, ncol=3, byrow=TRUE)
matA

```

## Assign Dimension Names

You can name the rows and columns of a matrix using the functions <code>dimnames()</code>, <code>rownames()</code> and <code>colnames()</code>:

```{r}
dim(matA)
dimnames(matA)
dimnames(matA) = list(c("row1","row2"),c("col1","col2","col3"))
matA
colnames(matA) = c("Col1", "Col2", "Col3")
rownames(matA) = c("Row1", "Row2")
matA
```

## Subset Elements of a Matrix

You can subset the elements of a matrix by position or name:

```{r}
# subset by position
matA[1, 2]
```

```{r}
# subset by name
matA["Row1", "Col1"]
```

You can extract entire rows or columns

```{r}
# extract first row
matA[1, ]
```

```{r}
# extract 2nd column
matA[, 2]
```

## Subset Elements of a Matrix

When you extract an entire row or column of a matrix, R drops the dimension attributes of the result. That is, what you get is a numeric vector and not a row or column matrix. To preserve the dimension attributes the optional argument <code>drop=FALSE</code>:

```{r}
# preserve dimension attributes using drop=FALSE
matA[1, , drop=FALSE]
matA[, 2, drop=FALSE]

```

## Create Vector Object

Vectors in R do not have dimension attributes. They appear as row vectors but that is just how they are printed on screen.

```{r}
xvec = c(1,2,3)
xvec
class(xvec)
dim(xvec)
names(xvec) = c("x1", "x2", "x3")
xvec

```

## Coerce Vector to Matrix

```{r}
# coerce vector to class matrix: note column vector is created
xvec = as.matrix(xvec)
xvec
class(xvec)
# create row vector
xvec = c(1,2,3)
xvec = matrix(xvec, nrow=1)
xvec
```

Note: R has many <code>as.</code> functions for coercing the class of an object to another class. 

## Transpose of Matrix

Transposing a matrix interchanges rows and columns. The transpose function in R is <code>t()</code>:

```{r}
matA = matrix(data=c(1,2,3,4,5,6), nrow=2, ncol=3, 
              byrow=TRUE)
t(matA)
xvec = c(1,2,3)
t(xvec)

```

## Symmetric Matrix

A symmetric matrix is equal to its transpose:

```{r}

matS = matrix(c(1,2,2,1),2,2)
matS
# check for symmetry
matS == t(matS)

```

## Basic Matrix Operations (element by element)

Element by element operations require that all matrices have the same dimension and apply to each element the matrices.

```{r}
matA = matrix(c(4,9,2,1),2,2, byrow=TRUE)
matB = matrix(c(2,0,0,7),2,2, byrow=TRUE)
matA
matB

```

## Matrix Addition and Subtraction

Matrix addition (<code>+</code>) and subtraction (<code>-</code>) are element by element operations:

```{r}
# matrix addition
matC = matA + matB
matC
# matrix subtraction
matC = matA - matB
matC

```

## Scalar Multiplication

Multipling a matrix by a scalar (single number) is an element by element operation:

```{r}
matA = matrix(c(3,-1,0,5), 2, 2, 
              byrow=TRUE)
matA
matC = 2*matA
matC

```

## Matrix Multiplication

Matrix multiplication only works on conformable matrices:

```{r}
matA = matrix(1:4,2,2, byrow=TRUE)
matB = matrix(c(1,2,1,3,4,2),2,3, byrow=TRUE)
matA
matB
dim(matA)
dim(matB)
```

## Matrix Multiplication (not element by element)

In R, <code>%*%</code> performas matrix multiplication: 

```{r}
matC = matA%*%matB
matC
```

The $(i,j)$ element of <code>matC</code> is the dot product of the i'th row of <code>matA</code> with the j'th column of <code>matB</code>

Multiplying two non-conformable matrices results in an error:

```{r}
# here B%*%A doesn't work b/c A and B are not comformable
# matB%*%matA
# Error in matB %*% matA : non-conformable arguments
```

## Identity Matrix

Create identity matrix of dimension 2 using the <code>diag()</code> function:

```{r}
matI = diag(2)
matI
```

Multiplying a matrix by the identity matrix returns the original matrix:

```{r}
matA = matrix(c(1,2,3,4), 2, 2, byrow=TRUE)
matI%*%matA
matA%*%matI

```

## Diagonal Matrices

Create a diagonal matrix with $c=(1,2,3)$ along the main diagonal.

```{r}
matD = diag(1:3)
matD
```

## Matrix Inverse

In R, the function <code>solve()</code> computes the inverse of a matrix:

```{r}
matA
matA.inv = solve(matA)
matA.inv
matA%*%matA.inv
matA.inv%*%matA

```


## Solving Systems of Linear Equations

The solution of two linear equations in two unknows is at the intersection of the two lines:

```{r}
curve(1-x, 0, 1, lwd=2, ylab="y")
abline(a=-1, b=2, lwd=2)
title("x+y=1, 2x-y=1")

```

## Solving Systems of Linear Equations

The system of linear equations can be expressed in matrix form as $Az = b$. The solution is $z = A^{-1}b$: 

```{r}
matA = matrix(c(1,1,2,-1), 2, 2, byrow=TRUE)
vecB = c(1,1)
matA.inv = solve(matA)
z = matA.inv%*%vecB
z

```

## Summation Using Matrix Algebra

$$ \sum_{k=1}^{n}x_{k} = \mathbf{x}^{\prime}\mathbf{1} $$

```{r}
onevec = rep(1,3)
xvec = c(1,2,3)
# sum elements in x - returns 1 x 1 matrix
t(xvec)%*%onevec
# more efficient
crossprod(xvec, onevec)
# more efficient still - returns a number
sum(xvec)

```

## Sum of Squares

$$ \sum_{k=1}^{n}x_{k}^{2} = \mathbf{x}^{\prime}\mathbf{x} $$

```{r}
xvec
t(xvec)%*%xvec
# more efficient
crossprod(xvec)
# most efficient
sum(xvec^2)

```

## Sum of Cross Products

$$ \sum_{k=1}^{n}x_{k}y_{k} = \mathbf{x}^{\prime}\mathbf{y} $$

```{r}
yvec = 4:6
xvec
yvec
t(xvec)%*%yvec
crossprod(yvec, xvec)
sum(xvec*yvec)
```

## Partitioned Matrices

```{r}
A11mat = matrix(c(1,2,5,6), 2, 2, byrow=TRUE)
A12mat = matrix(c(3,4,7,8), 2, 2, byrow=TRUE)
A21mat = matrix(c(9,10,13,14), 2, 2, byrow=TRUE)
A22mat = matrix(c(11,12,15,16), 2, 2, byrow=TRUE)
Amat = rbind(cbind(A11mat, A12mat), cbind(A21mat, A22mat))
Amat
```

## Rank of a matrix

```{r}
library(Matrix)
Amat = matrix(c(1,3,5,2,4,6), 2, 3, byrow=TRUE)
as.numeric(rankMatrix(Amat))
```

## Cholesky Factorization

```{r}
Sigma = matrix(c(1, 0.75, 0.75, 1), 2, 2)
Sigma
```

```{r}
C = chol(Sigma)
C
t(C)%*%C
```

Singular correlation matrix

```{r}
Sigma = matrix(c(1, 1, 1, 1), 2, 2)
Sigma
chol(Sigma, pivot=TRUE)
```

