% Turabian Formatting for Research Papers Template, 2016/10/09
%
% Developed using the turabian-formatting package (2016/10/09), available through CTAN: http://www.ctan.org/pkg/turabian-formatting
%
% Additional document class formatting options:
%
% raggedright: ragged right formatting without hyphenations
% authordate: support for the author-date citation style
% endnotes: support for endnotes



\documentclass{turabian-researchpaper}


\usepackage[utf8]{inputenc}
\usepackage{csquotes, ellipsis}

% Specify paper size with geometry package
\usepackage[pass, letterpaper]{geometry}

% For citations, use the biblatex-chicago package
\usepackage{biblatex-chicago}
\addbibresource{works-cited.bib}


% Information for title page
\title{Turabian-formatted Research Paper:}
\subtitle{A Template based on Turabian's \emph{A Manual for Writers}, 8th edition}
\author{Author's Name}
\course{Course}
\date{\today}


\begin{document}

\maketitle


\section{Introduction}

Amazing, introductory ideas that provide unique insight into your field of interest and ``wows" your professor---formatted based on Kate L. Turabian's \emph{A Manual for Writers of Research Papers, Theses, and Dissertations} (8th edition).


\section{An Interesting Section}

Great thoughts that further your argument. This includes lots of strong evidence presented throughout several paragraphs, each accompanied by necessary citations.
\begin{quotation}
    \noindent Here is a block quotation---a passage from a text you found insightful and wanted to share with others. Maybe it is from a journal article, website, or book. Irrespective, it should support the argument being made.\footnote{A citation for the quoted material.}
\end{quotation}
Maybe a sentence or two that bring the argument and evidence together.\autocite[34]{example_source}


\section{Another Insightful Section}

More ideas that really make this a great paper. Maybe a footnote or two.\footnote{Some peripheral thoughts that belong in a note.}


\section{Conclusions}

At this point, you've changed everything (including your marks!). Time to wrap up!


\clearpage
\printbibliography

\end{document}
